A Road Segment Prioritization Approach for Cycling Infrastructure

### 2023-04-24 README.tex
plot for paper

\title{A Road Segment Prioritization Approach for Cycling Infrastructure}
\author{true \and true \and true}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Understanding the motivators and deterrents to cycling is essential for creating infrastructure that gets more people to adopt cycling as a mode of transport. This paper demonstrates a new approach to support the prioritization of cycling infrastructure and cycling network design, accounting for cyclist preferences and the growing emphasis on `filtered permeability' and `Low Traffic Neighborhood' interventions internationally. The approach combines distance decay, route calculation, and network analysis methods to examine where future cycling demand is most likely to arise, how such demand could be accommodated within existing street networks, and how to ensure a fair distribution of investment. Although each of these methods has been applied to cycling infrastructure prioritization in previous research, this is the first time that they have been combined, creating an integrated road segment prioritization approach. The approach, which can be applied to other cities, is demonstrated in a case study of Manchester, resulting in cycling networks that balance directness against the need for safe and stress-free routes under different investment scenarios. A key benefit of the approach from a policy perspective is its ability to support egalitarian and cost-effective strategic cycle network planning.

\par

\textbf{Keywords:} cycling networks, low-traffic neighborhoods, routing, transport equity
\end{abstract}

@@

@@

Our work builds on these community finding approaches by proposing a
similar greedy network expansion algorithm for cycle network expansion
\emph{within communities}. We incorporate community finding methods for study
area partitioning with weighted routing to avoid links that are
stressful to cycle on. In doing so, we propose an approach that ensures a better distribution
investments across the study area while also accounting for motivators and deterrents to cycling.
We propose three sub-methods that address some of the limitations of previous
studies. These limitations include (a) an inherent bias when basing network
design solely on existing cycling demand, (b) the proposal of routes that may
not correspond to studies on cyclist preference and government policies,
and (c) an insufficient consideration of the ethical principles underlying
the analysis. The current work addresses these limitations through the following organisation. Section \ref{calculating-potential-cycling-demand}
focuses on calculating potential cycling demand. Section \ref{routing}
focuses on routing the demand onto the road network while accounting for
cyclist preferences and government priorities. Section
\ref{road-segment-prioritization} outlines a method for partitioning the
study area based on a community finding algorithm and routed cycling
demand. It then introduces the network expansion algorithms, and compares an approach grounded in
`egalitarianism' to one grounded in `utilitarianism'.

\hypertarget{calculating-potential-cycling-demand}{%
\section{Calculating Potential Cycling Demand}\label{calculating-potential-cycling-demand}}

@@

\emph{OD} data can be obtained from a range of sources,
including GPS data, household travels surveys and census data on work
locations. In areas where observed OD data is unavailable, modelling
techniques such as spatial interaction models can be used to estimate
travel volumes between zones (Black, 1995; Martínez and Viegas, 2013; Wilson, 1971). In this paper we use open access data from the UK
census (ONS, 2011), which contains aggregate statistics on
number of commuters between administrative zones --- Middle layer Super
Output Areas (MSOA) --- by mode of travel. MSOAs have an average
population of just over 8000 (ONS, 2018).\footnote{ See
  \url{https://wicid.ukdataservice.ac.uk/} for open access to the OD data.}
Figures \ref{fig:potdemhistograms} and \ref{fig:desirefacetcycling}
illustrate the proportion of trips cycled by distance and the
geographic extent of the input OD dataset used in this paper.

@@

\hypertarget{routing}{%
\section{Routing}\label{routing}}

@@

The above points are accounted for by using a weighted road network for
routing. This has been previously done by multiplying all road segments
without cycling infrastructure by a fixed impedance factor
(Mauttone et al., 2017), or by assigning different weights to the road
segments proportional to the investment cost of bringing them to an
acceptable level of stress for cycling (Duthie and Unnikrishnan, 2014).
Perceived stress for cyclists has been found to be a function of vehicular traffic
volume and speed (Sorton and Walsh, 1994), both of which vary predictably
with road type. Gehrke et al. (2020) build on this information to use
routing impedance factors for road segments that are based on road type
and the existence of cycling infrastructure.

For the purposes of this paper we created a similar weighting profile
that is adjusted to favor less stressful roads (based on the definition
by Sorton and Walsh (1994) and information from Table
\ref{table:osmroadtypes}) and roads with existing cycling
infrastructure. We believe this to be more appropriate than the simplistic
approach adopted by Mauttone et al. (2017) as it makes use of a hierarchy
of road preference based on perceived stress levels, going beyond simply
favouring roads with existing cycling infrastructure.

@@

\hypertarget{road-segment-prioritization}{%
\section{Road Segment Prioritization}\label{road-segment-prioritization}}

@@

\hypertarget{community-detection}{%
\subsection{Community Detection}\label{community-detection}}

@@

\hypertarget{algorithm-1-utilitarian-expansion}{%
\subsection{Algorithm 1: Utilitarian Expansion}\label{algorithm-1-utilitarian-expansion}}

@@

\hypertarget{algorithm-2-egalitarian-expansion}{%
\subsection{Algorithm 2: Egalitarian Expansion}\label{algorithm-2-egalitarian-expansion}}

@@

\hypertarget{connectivity}{%
\subsection{Connectivity}\label{connectivity}}

@@

The approach is not without limitations. The level of detail is only as
good as the granularity of the available data (in this case relatively
coarse zones). Iacono et al. (2010) note that such large travel zones
are not ideal for understanding route choice behaviour of cyclists and
pedestrians. They also give rise to an `ecological fallacy' whereby
average characteristics are assumed to apply to all residents of the
aggregated geographical area, suggesting a need for applying the methods
to more granular OD data (and for governments and other
data-collecting organisations to make OD data more
readily available).
The potential demand calculation is also based on cycling in the traditional sense, and does not consider the
effect of micro-mobility on reducing topology-related impedance to cycling. Given that we are proposing an approach which can accommodate any cycling uptake functions, this is an acceptable limitation.

@@

@@

>>>>

\title{A Road Segment Prioritization Approach for Cycling Infrastructure}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle
\begin{abstract}
Understanding the motivators and deterrents to cycling is essential for creating infrastructure that gets more people to adopt cycling as a mode of transport. This paper demonstrates a new approach to support the prioritization of cycling infrastructure and cycling network design, accounting for cyclist preferences and the growing emphasis on `filtered permeability' and `Low Traffic Neighborhood' interventions internationally. The approach combines distance decay, route calculation, and network analysis methods to examine where future cycling demand is most likely to arise, how such demand could be accommodated within existing street networks, and how to ensure a fair distribution of investment. Although each of these methods has been applied to cycling infrastructure prioritization in previous research, this is the first time that they have been combined, creating an integrated road segment prioritization approach. The approach, which can be applied to other cities, is demonstrated in a case study of Manchester, resulting in cycling networks that balance directness against the need for safe and stress-free routes under different investment scenarios. A key benefit of the approach from a policy perspective is its ability to support egalitarian and cost-effective strategic cycle network planning. ¶ \par ¶ \textbf{Keywords:} cycling networks, low-traffic neighborhoods, routing, transport equity
\end{abstract}

@@

\hypertarget{method-and-application}{%
\section{Method and Application}\label{method-and-application}}

\hypertarget{method-overview}{%
\subsection{Method overview}\label{method-overview}}

@@

Our work builds on these community finding approaches by proposing a
similar greedy network expansion algorithm for cycle network expansion
\emph{within communities}. We incorporate community finding methods for study
area partitioning with weighted routing to avoid links that are
stressful to cycle on. In doing so, we propose an approach that ensures a better distribution
investments across the study area while also accounting for motivators and deterrents to cycling.
We propose three sub-methods that address some of the limitations of previous
studies. These limitations include (a) an inherent bias when basing network
design solely on existing cycling demand, (b) the proposal of routes that may
not correspond to studies on cyclist preference and government policies,
and (c) an insufficient consideration of the ethical principles underlying
the analysis. The current work addresses these limitations through the following organisation. Section \ref{calculating-potential-cycling-demand}
focuses on calculating potential cycling demand. Section \ref{routing}
focuses on routing the demand onto the road network while accounting for
cyclist preferences and government priorities. Section
\ref{road-segment-prioritization} outlines a method for partitioning the
study area based on a community finding algorithm and routed cycling
demand. It then introduces the network expansion algorithms, and compares an approach grounded in
`egalitarianism' to one grounded in `utilitarianism'. As mentioned above, Manchester is used as a case study.
An overview of the steps involved is shown in Figure \ref{fig:methodflowchart}.

{\centering \includegraphics[width=0.8\linewidth]{extras/methodology_flowchart}

\caption{Flowchart of proposed methodology}\label{fig:methodflowchart}
\end{figure}

\hypertarget{calculating-potential-cycling-demand}{%
\subsection{Calculating Potential Cycling Demand}\label{calculating-potential-cycling-demand}}

@@

\emph{OD} data can be obtained from a range of sources,
including GPS data, household travels surveys and census data on work
locations. In areas where observed OD data is unavailable, modelling
techniques such as spatial interaction models can be used to estimate
travel volumes between zones (Black, 1995; Martínez and Viegas, 2013; Wilson, 1971). In this paper we use open access data from the UK
census (ONS, 2011), which contains aggregate statistics on
number of commuters between administrative zones --- Middle layer Super
Output Areas (MSOA) --- by mode of travel (this is similar to the approach
used by Grisé and El-Geneidy (2018)).
MSOAs have an average
population of just over 8000 (ONS, 2018).\footnote{ See
  \url{https://wicid.ukdataservice.ac.uk/} for open access to the OD data.}
Figures \ref{fig:potdemhistograms} and \ref{fig:desirefacetcycling}
illustrate the proportion of trips cycled by distance and the
geographic extent of the input OD dataset used in this paper.

@@

\hypertarget{routing}{%
\subsection{Routing}\label{routing}}

@@

The above points are accounted for by using a weighted road network for
routing. This has been previously done by multiplying all road segments
without cycling infrastructure by a fixed impedance factor
(Mauttone et al., 2017), by assigning different weights to the road
segments that are (a) proportional to the investment cost of bringing them to an
acceptable level of stress for cycling (Duthie and Unnikrishnan, 2014), or (b) inversely proportional to
perceived level of stress experienced by cyclists (Wysling and Purves, 2022).
Perceived stress for cyclists has been found to be a function of vehicular traffic
volume and speed (Sorton and Walsh, 1994), both of which vary predictably
with road type. Gehrke et al. (2020) build on this information to use
routing impedance factors for road segments that are based on road type
and the existence of cycling infrastructure.

For the purposes of this paper we created a similar weighting profile
that is adjusted to favor less stressful roads (based on the definition
by Sorton and Walsh (1994) and information from Table
\ref{table:osmroadtypes}) and roads with existing cycling
infrastructure. We believe this to be more appropriate than the simplistic
approach adopted by Mauttone et al. (2017) as it makes use of a hierarchy
of road preference based on perceived stress levels, going beyond simply
favouring roads with existing cycling infrastructure.
It is also different to the weighting profile used by Wysling and Purves (2022) as it adds a larger
weight on trunk and primary roads then other roads without cycling infrastructure.

@@

\hypertarget{road-segment-prioritization}{%
\subsection{Road Segment Prioritization}\label{road-segment-prioritization}}

@@

\hypertarget{community-detection}{%
\subsubsection{Community Detection}\label{community-detection}}

@@

\hypertarget{algorithm-1-utilitarian-expansion}{%
\subsubsection{Algorithm 1: Utilitarian Expansion}\label{algorithm-1-utilitarian-expansion}}

@@

\hypertarget{algorithm-2-egalitarian-expansion}{%
\subsubsection{Algorithm 2: Egalitarian Expansion}\label{algorithm-2-egalitarian-expansion}}

@@

\hypertarget{connectivity}{%
\subsubsection{Connectivity}\label{connectivity}}

@@

The approach is not without limitations. The level of detail is only as
good as the granularity of the available data (in this case relatively
coarse zones). Iacono et al. (2010) note that such large travel zones
are not ideal for understanding route choice behaviour of cyclists and
pedestrians. They also give rise to an `ecological fallacy' whereby
average characteristics are assumed to apply to all residents of the
aggregated geographical area, suggesting a need for applying the methods
to more granular OD data (and for governments and other
data-collecting organisations to make OD data more
readily available).
The potential demand calculation is also based on cycling in the traditional sense, and does not consider the
effect of micro-mobility on reducing topology-related impedance to cycling. Given that we are proposing an approach which can accommodate any cycling uptake functions, this is an acceptable limitation.
The network prioritization is based strongly on potential cycling demand calculations. While this provides a solid supplementary tool for network planning, it would be more useful as part of a multi-criteria approach that considers on-the ground information from existing cyclists, road geometries, and traffic analysis. Grisé and El-Geneidy (2018) propose a multi-criteria approach that includes the first of these points.

@@

\leavevmode\vadjust pre{\hypertarget{ref-grise2018if}{}}%
Grisé, E., El-Geneidy, A., 2018. If we build it, who will benefit? A multi-criteria approach for the prioritization of new bicycle lanes in quebec city, canada. Journal of transport and land use 11, 217--235.

@@

\leavevmode\vadjust pre{\hypertarget{ref-wysling2022improve}{}}%
Wysling, L., Purves, R.S., 2022. Where to improve cycling infrastructure? Assessing bicycle suitability and bikeability with open data in the city of paris. Transportation research interdisciplinary perspectives 15, 100648.
